---
title: "Data Checks and Cleaning Decisions (freMTPL2)"
format: gfm
execute:
  warning: false
  message: false
---

## Purpose

Check the consistency and distribution of freMTPL2 and decide how each variable is cleaned. The cleaning follows Noll et al.(2018) unless stated otherwise.

```{r}
#| message: false
library(here)
library(data.table)
library(ggplot2)

freq <- as.data.table(readRDS(here("data", "raw", "freMTPL2freq.rds")))
sev <- as.data.table(readRDS(here("data", "raw", "freMTPL2sev.rds")))

freq[, IDpol := as.integer(as.character(IDpol))]
sev[, IDpol := as.integer(as.character(IDpol))]

str(freq)
str(sev)
```

```{r}
# Check that ClaimNb in freq matches the number of claims recorded in sev
sev_count <- sev[, .(SevNb = .N), by = IDpol]

check <- merge(
  freq[, .(IDpol, ClaimNb)],
  sev_count,
  by = "IDpol",
  all.x = TRUE
)

check[is.na(SevNb), SevNb := 0]    # policies not in sev have no recorded claims

check[, consistent := ClaimNb == SevNb]

table(check$consistent)

# Claims in sev whose policy is missing from freq
sev[!freq, on = "IDpol", .N]
```

## 1. Frequency data

### 1.1 ClaimNb

```{r}
freq[, .N, by = ClaimNb][order(ClaimNb)]
```


```{r}
freq[ClaimNb >= 4, .(ClaimNb, Exposure, annual_rate = ClaimNb / Exposure)][order(-ClaimNb)]
```

**Decision**: Following previous studies, we cap ClaimNb at 4 to limit the effect of extreme claim counts.

### 1.2 Exposure

```{r}
freq[Exposure > 1, .N]
```

**Decision**: Following previous studies, we cap Exposure at 1.0. Values above 1 contradict to the definition.

### 1.3 VehPower

```{r}
freq[, .N, by = VehPower][order(VehPower)]
```

**Decision**: VehPower has some sparse high-power groups and also shows an implausible upward turn at the maximum. Following previous studies, we deal with it as a categorical variable and group values of 9 and above into a single group.

### 1.4 VehAge

```{r}
summary(freq$VehAge)

hist(
  freq$VehAge,
  breaks = seq(0, 100, by = 5),
  xlab = "VehAge",
  ylab = "N"
)
```

```{r}
freq[VehAge >= 80, .N, by = VehAge][order(VehAge)]
```

**Decision**: VehAge is grouped into classes and treated as a categorical variable, following previous studies. The open-ended top class absorbs the sparse and erratic observations at the upper end.

### 1.5 DrivAge

```{r}
summary(freq$DrivAge)

hist(
  freq$DrivAge,
  breaks = seq(0, 100, by = 5),
  xlab = "DrivAge",
  ylab = "N"
)
```

```{r}
freq[DrivAge >= 91, .N, by = DrivAge][order(DrivAge)]
```

**Decision**: DrivAge is grouped into classes and treated as a categorical variable, following previous studies. The open-ended top class absorbs the sparse and erratic observations at 99.

### 1.6 BonusMalus

```{r}
summary(freq$BonusMalus)

hist(
  freq$BonusMalus,
  breaks = seq(50, 230, by = 10),
  xlab = "BonusMalus",
  ylab = "N"
)
```

**Decision**: We cap it at 150 by following previous studies.

### 1.7 Area vs Density

```{r}
summary(freq$Area)
summary(freq$Density)
cor(as.integer(freq$Area), log(freq$Density))
```

**Decision**: We use only log Density.

## 2. Severity data

```{r}
# Share of the total claim amount from the largest 1, 5, 10 and 50 claims
sev[order(-ClaimAmount),
  .(top_k = c(1, 5, 10, 50),
    share = cumsum(ClaimAmount)[c(1, 5, 10, 50)] / sum(ClaimAmount))]

sev[order(-ClaimAmount)][1:10]
```

```{r}
# For each candidate cap u:
#   pct_excess: share of the total amount above u
#   load_L    : flat load that restores the uncapped total
thr <- c(1e4, 2e4, 5e4, 1e5, 2e5, 5e5)

rbindlist(lapply(thr, function(u) sev[, .(
  u          = u,
  n_above    = sum(ClaimAmount > u),
  pct_count  = mean(ClaimAmount > u),
  pct_excess = sum(pmax(ClaimAmount - u, 0)) / sum(ClaimAmount),
  load_L     = sum(ClaimAmount) / sum(pmin(ClaimAmount, u))
)]))
```

**Decision**: Claim amounts are capped at 100k per claim. This affects only 41 claims (0.16%) and limits the influence of extreme losses on the Gamma GLM. The excess (17.7% of the amount) is added back through a flat load L ≈ 1.22 on the full data; the load used in the model is re-estimated on the training data.

## References

Noll, A., Salzmann, R. and Wüthrich, M. V. (2018).
[Case Study: French Motor Third-Party Liability Claims](https://papers.ssrn.com/sol3/papers.cfm?abstract_id=3164764).
SSRN.

## Session Info

```{r}
sessionInfo()
```