---
title: "Pure Premium Modelling with Frequency-Severity GLMs (freMTPL2)"
format: gfm
execute:
  warning: false
  message: false
---

## Summary
 
- **Frequency:** A Poisson GLM beat the intercept-only baseline by 4.4% in
  test deviance, mainly by capturing the high claim rates of young drivers.
- **Severity:** Gamma GLMs overfit to a few large claims and did worse than
  a constant mean on the test set, so the baseline is the better choice here.
- **Pure premium:** Frequency GLM × constant severity ranked risks best on
  the test set (Gini 0.31, top-to-bottom decile lift 7.3); adding a severity
  GLM did not help.
- **Calibration:** All models match the total claim level on the training
  data (A/E ≈ 1); the test deviations of 2–4% are within sampling noise.
  
## Data
 
We use the French motor third-party liability data freMTPL2 from
CASdatasets. The cleaning decisions are documented in `02_data_check.qmd`.
Policies are randomly split 80/20 into train and test sets, and all claims
of a policy go to the same set.

```{r}
#| message: false
library(here)
library(data.table)
library(ggplot2)

source(here("R", "preprocess.R"))
source(here("R", "split.R"))
source(here("R", "metrics.R"))

freq <- as.data.table(readRDS(here("data", "raw", "freMTPL2freq.rds")))
sev <- as.data.table(readRDS(here("data", "raw", "freMTPL2sev.rds")))

dat <- do.call(split_data, prepare_data(freq, sev))

summary(dat$freq_train)
summary(dat$sev_train)
```

## Frequency
 
We fit a Poisson GLM with log exposure as an offset and compare it with an
intercept-only model on the test set.

```{r}
freq_m0 <- glm(ClaimNb ~ 1 + offset(log(Exposure)),
               family = poisson(), data = dat$freq_train)

freq_m1 <- glm(
  ClaimNb ~ VehPowerGrp + VehAgeGrp + DrivAgeGrp + BonusMalus +
    VehBrand + VehGas + LogDensity + Region + offset(log(Exposure)),
  family = poisson(),
  data = dat$freq_train
)

freq_models <- list(null = freq_m0, glm = freq_m1)

freq_result <- rbindlist(lapply(freq_models, function(m) {
  mu <- predict(m, dat$freq_test, type = "response")
  freq_metrics(dat$freq_test$ClaimNb, mu)
}), idcol = "model")

knitr::kable(freq_result, digits = 4)
```

The GLM reduces test deviance by about 4.4% relative to the intercept-only
model.

### Actual vs predicted by driver age
 
Points show the observed and lines the predicted average frequency. Ages 85
and above are pooled into 85.

```{r}
freq_by_age <- function(models, data) {
  d <- data[, .(DrivAge = pmin(DrivAge, 85), Exposure, actual = ClaimNb)]
  for (nm in names(models)) d[, (nm) := predict(models[[nm]], data, type = "response")]
  d <- melt(d, id.vars = c("DrivAge", "Exposure"), variable.name = "series", value.name = "y")
  d[, .(freq = sum(y) / sum(Exposure)), by = .(DrivAge, series)]
}

freq_plot <- rbindlist(list(
  train = freq_by_age(freq_models, dat$freq_train),
  test = freq_by_age(freq_models, dat$freq_test)
), idcol = "data")
freq_plot[, data := factor(data, c("train", "test"))]

ggplot(mapping = aes(DrivAge, freq)) +
  geom_point(data = freq_plot[series == "actual"], colour = "grey40") +
  geom_line(data = freq_plot[series != "actual"], aes(colour = series)) +
  facet_wrap(~ data) +
  labs(x = "DrivAge", y = "average frequency", colour = "model")
```

The GLM captures the high claim frequency of young drivers.

## Severity

We fit Gamma GLMs with a log link on the capped claim amounts: a full model
with the same variables as the frequency model, and a smaller model without
VehBrand and Region.

```{r}
sev_m0 <- glm(ClaimAmountCap ~ 1, family = Gamma(link = "log"), data = dat$sev_train)

sev_m1 <- glm(
  ClaimAmountCap ~ VehPowerGrp + VehAgeGrp + DrivAgeGrp + BonusMalus +
    VehBrand + VehGas + LogDensity + Region,
  family = Gamma(link = "log"), data = dat$sev_train
)

sev_m2 <- glm(
  ClaimAmountCap ~ VehPowerGrp + VehAgeGrp + DrivAgeGrp + BonusMalus +
    VehGas + LogDensity,
  family = Gamma(link = "log"), data = dat$sev_train
)

sev_models <- list(null = sev_m0, glm = sev_m1, glm_small = sev_m2)

eval_sev <- function(models, data) {
  rbindlist(lapply(models, function(m) {
    mu <- predict(m, data, type = "response")
    sev_metrics(data$ClaimAmountCap, mu)
  }), idcol = "model")
}

sev_result <- rbindlist(list(
  train = eval_sev(sev_models, dat$sev_train),
  test = eval_sev(sev_models, dat$sev_test)
), idcol = "data")

knitr::kable(sev_result, digits = 4)
```

Both GLMs improve the training deviance but do worse than the
intercept-only model on the test set, which indicates overfitting.

All models have a test A/E below 1, because the mean severity in the test
set is lower than in the training set. We compare this gap with the
standard error of the mean test severity.
 
```{r}
x <- dat$sev_test$ClaimAmountCap
se_rel <- sd(x) / mean(x) / sqrt(length(x))
ae_test <- sev_result[data == "test" & model == "null", ae]
 
c(se_rel = se_rel, n_se = (1 - ae_test) / se_rel)
```

The test A/E of the intercept-only model (`r round(ae_test, 3)`) differs
from 1 by `r round((1 - ae_test) / se_rel, 1)` standard errors, which is
consistent with sampling noise.

### Actual vs predicted by driver age

```{r}
sev_by_age <- function(models, data) {
  d <- data[, .(DrivAge = pmin(DrivAge, 85), actual = ClaimAmountCap)]
  for (nm in names(models)) d[, (nm) := predict(models[[nm]], data, type = "response")]
  d <- melt(d, id.vars = "DrivAge", variable.name = "series", value.name = "y")
  d[, .(sev = mean(y)), by = .(DrivAge, series)]
}

sev_plot <- rbindlist(list(
  train = sev_by_age(sev_models, dat$sev_train),
  test = sev_by_age(sev_models, dat$sev_test)
), idcol = "data")
sev_plot[, data := factor(data, c("train", "test"))]

ggplot(mapping = aes(DrivAge, sev)) +
  geom_point(data = sev_plot[series == "actual"], colour = "grey40") +
  geom_line(data = sev_plot[series != "actual"], aes(colour = series)) +
  facet_wrap(~ data) +
  labs(x = "DrivAge", y = "average severity", colour = "model")
```

The GLMs predict higher severity for drivers aged 18–20 and 71+, following
a few large claims in the training data.

## Pure premium
 
The predicted loss of a policy is its predicted claim count times its
predicted severity. We compare four combinations of frequency and severity
models on the test set, with the capped claim amounts as actual loss.

```{r}
pp_test <- copy(dat$freq_test)

loss_by_pol <- dat$sev_test[, .(loss = sum(ClaimAmountCap)), by = IDpol]
pp_test[loss_by_pol, loss := i.loss, on = "IDpol"]
pp_test[is.na(loss), loss := 0]
```

```{r}
pp_test[, `:=`(
  n_null = predict(freq_m0, pp_test, type = "response"),
  n_glm = predict(freq_m1, pp_test, type = "response"),
  s_null = predict(sev_m0, pp_test, type = "response"),
  s_glm = predict(sev_m1, pp_test, type = "response"),
  s_glm_small = predict(sev_m2, pp_test, type = "response")
)]

pp_preds <- list(
  "null x null" = pp_test$n_null * pp_test$s_null,
  "glm x null" = pp_test$n_glm * pp_test$s_null,
  "glm x glm" = pp_test$n_glm * pp_test$s_glm,
  "glm x glm_small" = pp_test$n_glm * pp_test$s_glm_small
)
```

```{r}
pp_result <- rbindlist(lapply(pp_preds, function(p) {
  pp_metrics(pp_test$loss, p, pp_test$Exposure)
}), idcol = "model")

knitr::kable(pp_result, digits = 3)
```

Frequency GLM × constant severity gives the highest lift and Gini. Adding a
severity GLM does not improve the risk ranking. The differences between the
GLM combinations are small; Gini is the more stable measure, since lift
depends only on the two extreme deciles.
 
In the lift chart, policies are grouped into ten deciles of equal exposure
by predicted pure premium. Points show the actual and lines the predicted
pure premium.

```{r}
#| label: lift-chart

lift_plot <- rbindlist(lapply(pp_preds[-1], function(p) {
  lift_table(pp_test$loss, p, pp_test$Exposure)
}), idcol = "model")

ggplot(lift_plot, aes(bin, actual_pp)) +
  geom_point(colour = "grey40") +
  geom_line(aes(y = pred_pp), colour = "steelblue") +
  facet_wrap(~ model) +
  scale_x_continuous(breaks = 1:10) +
  labs(x = "risk decile (predicted pure premium)", y = "pure premium per exposure")
```

### Large-loss load

```{r}
sev_load <- dat$sev_train[, sum(ClaimAmount) / sum(ClaimAmountCap)]
sev_load
```

Claims are capped at 100k for modelling. The final pure premium is the
capped prediction multiplied by a flat large-loss load of 1.15, estimated
on the training data only. Because the load is a constant multiplier, it
does not change the lift or Gini results above.
  
## Pricing implications

- **Differentiate on frequency.** The frequency GLM drives almost all of the
  risk ranking, so rating factors should be chosen and validated on
  claim frequency.
- **Use a single average severity.** A constant severity plus a flat
  large-loss load is sufficient. A severity GLM would add complexity and
  charge some groups more on the basis of a few large claims, without
  improving predictions.
- **Driver age matters.** Drivers aged 18–20 claim about three times as
  often as the portfolio average, which supports age-based premiums.

## Session info

```{r}
sessionInfo()
```
  
  